Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-02-01 paper/zhang-cook-laa-langrene-menendez.Rmd
finish proofreading

```{r tour-path, out.height="25%",out.width="50%", fig.cap="Each square (frame) represents the projected data with a corresponding basis. Blue frames are found by an optimisation algorithm iteratively whilst the white frames are constructed between two blue frames by geodesic interpolation."}
include_graphics(here::here("img/tour_path.png"))
```

@@

\noindent where $I_d$ is the $d$-dimensional identity matrix and the constraint requires the projection bases $\mathbf{A}$ to be orhtonormal.

Several features of this optimisation are worth-noticing. 1) The optimisation is constrained and the space of the bases is a d-dimensional sphere. 2) The objective function may not be differentiable for a constructed index function. 3) Although finding the global maximum is the goal of an optimisation problem, it is also interesting to inspect local maximum in projection pursuit since it could present unexpected interesting projections. 4) The optimisation procedure needs to be easy to compute since the tour animation needs to be played in real-time during the optimisation.

@@

To be able to make diagnostics on the optimisers, the algorithms need to populate a data structure with key elements of the algorithm. When the algorithms run, key information regarding the decision variable, objective function and hyper-parameters needs to be recorded and stored as a data object so that it is ready to be supplied to the plotting functions for diagnostics.

@@

Three main elements are recorded for the projection pursuit optimisers: 1) projection bases: $\mathbf{A}$, 2) index values: $I$, and 3) State: $S$, which labels the observation with detailed stage in the optimisation. For optimiser SA and SAJO, possible values of the state include `random_search`,  `new_basis`, and `interpolation`. pseudo derivative search has a wider variety of state including `new_basis`, `direction_search`, `best_direction_search`, `best_line_search`, and `interpolation`.

Multiple iterators are also  needed to index the data collected at different levels: $t$ being a unique identifier that prescribes the natural ordering of each observation while $j$ and $l$ being the counter of the outer and inner loop, respectively, in Algorithm \ref{random-search}, \ref{simulated-annealing} and \ref{search-geodesic} above. Other parameters of interest recorded include $V_1 = \text{method}$ that tags the name of the optimiser, and $V_2 = \text{alpha}$ that indicates the sampling neighbourhood size.  A matrix notation of the data structure is presented in Equation \ref{eq:data-structure}.

@@

\noindent where $T^{\prime} = T + k_{J}+ l_{J+1}$. Note that there is no output in iteration $J + 1$ since the optimiser cannot find a better basis in the last iteration and the algorithm terminates. The final basis found is $A_T$ with the highest index value $I_T$.

@@

The construction of diagnostic plots adopts the core concept in ggplot2:  grammar of graphics [@wickham2010layered]. In grammar of graphics, plots are not produced via calling the commands, named by the appearance of the plot, i.e., boxplot and histogram, but via the concept of stacked layers. Seeing plots as stacked layers empowers us to composite diagnostic plots with an emphasis on any variable in the data object without the redundancy of creating different commands for the same type of plots that highlights on different variables.

@@

A starting point of diagnosing an optimiser is to understand how many search the optimiser has conducted. One may want to simply plot the index value of the search points across its natural order. A point geometry may work well if each iteration has a similar number of points,  however, the plot will be disproportionate if some iterations have considerably more points than other - it over-emphasizes the iterations that have more search points! An alternative is to summarise the search in each iteration using boxplot where each iteration will be spaced out equally. Occasionally, one may still want to switch back to point geometry if the number of points is small and this can be achieved via the `cutoff` argument. Additional annotation can be added with new layers, for example, the number of points searched in each iteration can be added as text label at the bottom of each iteration; the anchor bases to interpolate are connected and highlighted in a larger size; and the colour of the last iteration is in a grey scale to indicate no better basis found in this iteration.

Figure \ref{fig:toy-search} shows the searches of two different optimisers: SA (left) and SAJO (right). Both optimisers quickly find better bases at the first few iterations,  then take longer in the later iterations, and finally finish when the maximum number of evaluation is reached. The anchor bases, the ones with the highest index value in each iteration, always have an increased index value in the optimiser SA while this is not the case for SAJO. This feature gives SA an advantage in simple example, like the one presented while the jump out component in SAJO allows for a more holistic search in more complicated examples.

@@

Another interest of diagnosing is to examine how the index value progresses for the points on the interpolation path since the projection on these bases are played by the tour animation. Trace plots are created with plotting the index value of the interpolation points against time. Figure \ref{fig:toy-interp} presents the trace plot of the two optimisation routines as \ref{fig:toy-search}. The trace plot is smooth in both plots and this is a character of the holes index function used. An interesting discovery is that with optimiser SAJO, the interpolator, in iteration 6-8, first passes through some bases with higher index value than the anchor bases.

@@

```{r toy-pca, fig.height=5, fig.cap = "The PCA plot of 1D projection problem on the 5-variable dataset, \\code{boa5}, using the holes indexwith two optimisers: SA and PD. All the bases in PD has been flipped and a grey dashed line has been annotated to indicate the symmetry of two starting bases."}
```

Apart from checking the progression of an optimiser, looking at how the points evaluated by the optimisers look like in the space is another interest. Given the orthonormality constraint, the space of projection bases $\mathbf{A}_{p \times d}$ is a $p \times d$ dimension sphere and  dimension reduction methods, i.e. principal component analysis is applied first to project all the bases onto a 2D space. In a projection pursuit guided tour optimisation, there are various type of bases involved: 1) The start basis; 2) The anchor bases that have the highest index value in each iteration; 3) The search bases that an optimiser evaluated to produce the anchor bases; 4) The interpolating bases on the interpolation path; and finally 5) the end basis. The importance of these basis differs with the most important ones being the start, interpolating and end bases. Anchor and search bases can be turned on with argument `details = TRUE`. Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign. This happens because the projection is invariant to the sign difference of the bases and so does the index value, however, this creates difficulties for comparing end bases. A flip sign device is implemented to flip the sign of all the bases in one optimisation if the situation described above happens and to ensure different routines finish close to each other.

Several annotations have been made to help understanding this plot. The original space of the bases is a high dimensional sphere and random bases on the sphere can be generated via the CRAN package \CRANpkg{geozoo} [@geozoo]. Along with the bases recorded during the optimisation and a zero basis, PCA is performed to get the first two PC coordinates of all the bases. The search space in the 2D space is a circle with the origin being the PC coordinates of the zero matrix, and radius estimated as the largest distance between the origin and all the bases. The theoretical best basis is known with simulated data and can be labelled for comparing the how close the ending basis to the theoretical one for different optimisers. Various aesthetics, i.e. size, alpha and colour, are applicable to emphasize the crucial elements and adjust for the presentation. For example, anchor points and search points are less important and a smaller size and alpha is used; The interpolation paths are more informative if alpha increases from start to finish (that is, the path is more opaque towards the end).

Figure \ref{fig:toy-pca} presents the PCA plot of a 1D projection of a 5-variable dataset, \code{boa5}, using the holes index with two optimisers, PD and SA, being compared. Both optimisers get close to the theoretical bases, annotated by "*" and in this particular example pseudo derivative search gets closer. One can also appreciate the nature of the two different searches: PD first evaluates points in a small neighbourhood for a promising direction, while SA evaluates points randomly in the search space to search for the next target. There are dashed lines annotated for SA and it describes the interruption of the interpolation, which has been briefly mentioned in Figure \ref{fig:toy-interp} and will be discussed in details in Section \ref{monotonic}.

@@

Animation is another display to show how the search progresses from start to finish in the space and an  `animate = TRUE` argument is used to enable the animation. Figure \ref{fig:toy-pca-animated} shows six frames from the animation of the PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information one can learn from this animation is that the SA search finds its end basis quicker than the PD search since SA has finished in the 5th frame while PD is still making more progression.

@@

As mentioned in Section \ref{toy-pca}, the original $p \times d$ dimension space can be simulated via randomly generated bases via \CRANpkg{geozoo} [@geozoo] package. A tour plot rotates these random bases and the ones collected by the optimisation in the original space and hence provides a stereoscopic view of the search. While the PCA plot projects the bases from the direction that maximises the variance, the tour plot display of the original high dimensional space from various directions using animation. Figure \ref{fig:toy-tour} shows the tour plot of the same two optimisations in the 5D space.

@@

In this section, several examples will be presented to show how the diagnostic plots discover something unexpected in the optimisation algorithm, and guide the implementation of some new features.

@@

Random variables with different distributions have been simulated and the distributional form of each variable is presented in Equations \ref{eq:sim-norm} to \ref{eq:sim-x7}. Variable `x1`, `x8`, `x9` and `x10` are normal noise with zero mean and unit variance, and `x2` to `x7` are normal mixtures with varied weights and locations. All the variables have been scaled to have an overall unit variance before projection pursuit.

@@

An example of non-monotonic interpolation has been given in Figure \ref{fig:toy-interp}: a path that passes bases with higher index value than the target one. With optimiser SAJO, a non-monotonic interpolation is justified since the target bases do not necessarily have a higher index value than the current one, while this is not the case for SA. The original trace plot for a 2D projection, optimised by SA, is shown on the left panel of Figure \ref{fig:interruption} and one can observe clearly that the non-monotonic interpolation has undermined the optimiser to realise its full potential. Hence, an interruption is implemented to stop the interpolation at the basis with the highest index value. The right panel of Figure \ref{fig:interruption} presents the trace plot after implementing such interruption and it can be observed that while the first two interpolations are identical, in the third interpolation the basis at time 61 has a higher index value than the target. Rather than starting the next iteration from the target basis on Time 65 as shown in the left panel, the optimiser, on the right panel, starts the next iteration from the interpolation basis on Time 61. With the interruption, the optimiser is able to find the final basis with a index value higher than the most promising basis when the interruption is not implemented.

@@

Once the final basis has been found by an optimiser, one may want to push further in the close neighbourhood to see if an even better basis can be found. A polish search takes the final basis of an optimiser as the start of a new guided tour to search for local breakthrough. The polish algorithm is similar to the SA with three main distinctions: 1) 100 rather than one candidate bases are generated each time in the inner loop; 2) the search neighbourhood, `alpha`, is reduced in the inner loop, rather than in the outer loop, once a set of candidate bases fails to make an improvement; and 3) three more termination conditions have been added to ensure the new basis generated is distinguishable from the current in terms of the distance in the space,  percentage change in the index value, and neighbourhood:

@@

Figure \ref{fig:polish} presents the trace plot of a 2D projection, optimised by SA and followed by the polish search, along with the projection on the final basis of each algorithm. The end basis found by SA reveals the four clusters in the data, but the edges of each cluster are not clean-cut. Polish works with this end basis and further pushes the index value to produce even clear structure in the data. One can observe that the projection after polishing better distinguishes the four clusters with more trimmed edges, espeically along the vertical axis.

@@

The index function before this section are all smooth, while this is not the case for all. A 1D projection function based on the Kolmogorov test, $I^{nk}(n)$, compares the projected data, $\mathbf{Y}_{n \times 1}$, to a randomly generated normal distribution, $\mathcal{N}_{n \times 1}$, based on the empirical cumulated distribution function (ECDF). Let $F_{.}(n)$ be the ECDF function, with two possible subscripts $Y$ and $\mathcal{N}$ representing the projected and randomly generated data, and $n$ denoting the number of observation, the Normal Kolmogorov, $I^{nk}(n)$, index is defined as:

$$I^{nk}(n) = \max \left[F_{Y}(n) - F_{\mathcal{N}}(n)\right]$$
With a non-smooth index function, two research questions are raised to understand

1) if any optimiser fails to optimise a non-smooth index; and 
2) whether the optimisers can find the global optimum despite the presence of a local optimum

```{r noisy-better-geo, fig.height = 6, fig.cap = "Trace and PCA plot of 1D projection on 5\\-variable dataset, \\code{boa5}, with \\code{norm\\_kol} index, optimised by all three optimiers. Optimiser PD fails to optimise the non-smooth index while although both SA and SAJO success in optimise the index, SAJO takes much longer than SA and finishes off closer to the theoretial best."}
```

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of all three optimisers. As expected, none of the interpolated path is smooth. There is barely any improvement made by PD, indicating its failure in optimising non-smooth index. While SA and SAJO have managed to get close to the index value of the theoretical best, trace plot shows that it takes SAJO much longer to interpolate towards the final basis. This long interpolation path is partially due to the fluctuation in the early iterations, where the SAJO tends to generously accept inferior basis before concentrating near the optimum. The PCA plot shows the interpolation path and search points excluding the last termination iteration. [A sentence for PD]. Comparing the amount of random search done by SA and SAJO, it is surprising that although SAJO doesn't actually carry that many of search as SA. Combine the insight from both the trace and PCA plot, one can learn the two different search strategies by SA and SAJO: SA frequently samples in the space and only make a move when an improvement will be made, while SAJO first broadly accepts basis in the space and only starts the extensive search in a narrowed subspace. The specific decision of which optimiser to use will depends on the index curve in the basis space but if the basis space is highly non-smooth and spike, accepting inferior bases at first, as SAJO, can lead to a more efficient search, in terms of the overall number of points evaluated.

```{r kol-result, fig.height = 5, fig.cap="PCA plot of 1D projection on the 6\\-variable data, \\code{boa6}, with normal kolmogorov index. Two optimisers: SA and SAJO and two search sizes: 0.5 and 0.7 are used in the simulation for 20 replicates. Local optimum (x), is presented in this experiment, along with the global optimum (*). The final basis is larger and more opaque than the start basis for each path and color represents whether the global or local optimum is found, after polishing."}
```

The next experiment compares the performance of SA and SAJO when a local maximum exists. Two search neighbourhood sizes: 0.5 and 0.7, are compared to understand how a large search neighbourhood would affect the final basis and the length of the search. Figure \ref{fig:kol-result} shows 80 paths simulated using 20 seeds in the PCA plot, faceted by  optimiser and search size. With SA and a search size of 0.5, despite being the simplest and fastest, SA fails to optimise in three instances where they finish neither near the local nor the global optimum. With a larger search size of 0.7, none of the seed fail and more seeds find the global maximum. Comparing SA and SAJO for a search size of 0.5, SAJO doesn't seem to improve the final basis found, despite having longer interpolation paths. However, the denser path near the local maximum is an indicator that SAJO is working hard to examine if there is any other optimum in the space but the relative small search size has diminished its ability to reach out to the true global maximum. With a larger search size, almost all the seeds (16 out of 20) have found the global maximum and some final bases are much closer to the theoretical best, as compared to the three other cases. This indicates that SAJO, with a reasonable large search size, is able to overcome the local optimum and optimise close towards the underlying true optimum.

@@

The implementation of this projection has been divided into two packages: the data collection object is implemented in the existing CRAN package \CRANpkg{tourr} [@tourr] while the optimiser diagnostics have been implemented in a new package, \pkg{ferrn}. When a guided tour is run, the users can choose if the data from optimisation should be collected via the `verbose` argument. Once the data object has been obtained, the package, \pkg{ferrn}, can provide four diagnostic plots as shown in Section \ref{vis-diag}. The structure of package functionality has been listed below.

@@

  - \code{explore\_trace\_search()} produces summary plots in Figure \ref{fig:toy-search}
  - \code{explore\_trace\_interp()}produces trace plots for the interpolation points in Figure \ref{fig:toy-interp}
  - \code{explore\_space\_pca()} produces the PCA plot of projection bases on the reduced space in Figure \ref{fig:toy-pca}. Animated version in Figure \ref{fig:toy-pca-animated} can be turned on via the argument `animate = TRUE`
  - \code{explore\_space\_tour()} produces animated tour view on the full space of the projection bases in Figure \ref{fig:toy-tour}.

@@

- Utilities
  - \code{add\_*()} are internal wrapper functions that facilitate the composition of PCA plot
  - \code{theme\_fern()} and \code{format\_label()} for better display of the grid lines and axis formatting
  - \code{clean\_method()} for clean up the name of the optimisers
  - \code{botanical\_palettes} is a collection of colour palettes from Australian native plants Quantitative palettes include daisy, banksia and cherry and sequential palettes contain fern and acacia
  - \code{botanical\_pal()} as the colour interpolator 
  - \code{scale\_color\_*()} and \code{scale\_fill\_*()} for scaling the colour and fill of the plot

>>>>

```{r tour-path, out.height="20%",out.width="50%", fig.cap="Each square (frame) represents the projected data with a corresponding basis. Blue frames are found by an optimisation algorithm iteratively whilst the white frames are constructed between two blue frames by geodesic interpolation."}
include_graphics(here::here("img/tour_path.png"))
```

@@

\noindent where $I_d$ is the $d$-dimensional identity matrix and the constraint requires the projection bases $\mathbf{A}$ to be orhtonormal. Several features of this optimisation are worth-noticing. 1) The optimisation is constrained and the space of the bases is a d-dimensional sphere. 2) The objective function may not be differentiable for a constructed index function. 3) Although finding the global maximum is the goal of an optimisation problem, it is also interesting to inspect local maximum in projection pursuit since it could present unexpected interesting projections. 4) The optimisation procedure needs to be easy to compute since the tour animation needs to be played in real-time during the optimisation.

@@

Before making diagnostic plots, information generated by the optimisation need to be collected and a pre-defined data structure solidifies the information being collected and streamlines the subsequent plot-making.

@@

Three main pieces of information are recorded for the projection pursuit optimisers: 1) projection bases $\mathbf{A}$, 2) index values $I$, and 3) state $S$. For SA and SAJO, possible state include `random_search`,  `new_basis`, and `interpolation`. Pseudo derivative (PD) has a wider variety of state including `new_basis`, `direction_search`, `best_direction_search`, `best_line_search`, and `interpolation`. Multiple iterators index the information collected at different levels: $t$ is a unique identifier prescribing the natural ordering of each observation;  $j$ and $l$ are the counter of the outer and inner loop respectively. Other parameters of interest recorded include `method` that tags the name of the optimiser, and `alpha` that indicates the sampling neighbourhood size for searching observations.  A matrix notation of the data structure is presented in Equation \ref{eq:data-structure}.

@@

\noindent where $T^{\prime} = T + k_{J}+ l_{J+1}$. Note that there is no output in iteration $J + 1$ since the optimiser doesn't find a better basis in the last iteration and terminates. The final basis found is $A_T$ with index value $I_T$.

@@

The construction of diagnostic plots adopts the core concept of grammar of graphics [@wickham2010layered] in ggplot2. In grammar of graphics, plots are not produced by calling the commands, named by the appearance of the plot, i.e., boxplot and histogram, but by the concept of stacked layers. Seeing plots as stacked layers gives analysts the freedom to composite plots with multiple elements on hands.

@@

A starting point of diagnosing an optimiser is to understand how many search an optimiser has conducted. One may want to simply plot the index value of the search points across its natural order, but a point geometry may work well if each iteration has a similar number of points. When some iterations have considerably more points than others, using a point geometry over-emphasizes the iterations that have more search points since these iterations will occupy the mast majority of the plot space. An alternative is to summarise the search in each iteration using boxplot and each iteration will then be spaced out equally. Occasionally, one may still want to switch back to a point geometry if the number of points is small in a particular iteration and this can be achieved via the `cutoff` argument in the search plot function. Additional annotations are added to facilitate better reading of the plot and these includes  
1) the number of points searched in each iteration can be added as text label at the bottom of each iteration; 2) the anchor bases to interpolate are connected and highlighted in a larger size; and 3) the colour of the last iteration is in a grey scale to indicate no better basis found in this iteration.

Figure \ref{fig:toy-search} shows an example of the search plot for SA (left) and SAJO (right). Both optimisers quickly find better bases in the first few iterations and then take longer to find one in the later iterations. The anchor bases, the ones found with the highest index value in each iteration, always have an increased index value in the optimiser SA while this is not the case for SAJO. This feature gives SA an advantage in this simple example to quickly find the optimum.

@@

Another interest of diagnostic is to examine how the index value improves for interpolating bases since the projection on these bases is played by the tour animation. Trace plots are created with plotting the index value of the interpolation points against time. Figure \ref{fig:toy-interp} presents the trace plot of the same optimisation routine as Figure \ref{fig:toy-search}. The trace plot is smooth and an interesting discovery is that in SAJO, the interpolation in iteration 6-8 is not monotonic. Sometimes, this non-monotonicity is acceptable since the target basis does not necessarily have a higher index value than the current one, while in some cases, the optimisation will be better off if these intermediate basis can be used as the start of the next iteration (see section \ref{monotonic} for a detailed discussion on this).

@@

```{r toy-pca, fig.cap = "The PCA plot of 1D projection problem on the 5-variable dataset, \\code{boa5}, using the holes index with two optimisers: SA and PD. All the bases in PD has been flipped and a grey dashed line has been annotated to indicate the symmetry of two starting bases."}
```

Apart from checking the search and progression of an optimiser, looking at where the bases are positioned in the basis space is another interest. Given the orthonormality constraint, the space of projection bases $\mathbf{A}_{p \times d}$ is a $p \times d$ dimension sphere and  dimension reduction methods, i.e. principal component analysis is applied to first project all the bases onto a 2D space. In a projection pursuit guided tour optimisation, there are various type of bases involved: 1) The start basis; 2) The search bases that an optimiser evaluated to produce the anchor bases; 3) The anchor bases that have the highest index value in each iteration; 4) The interpolating bases on the interpolation path; and finally 5) the end basis. The importance of these basis differs and the most important ones are the start, interpolating and end bases. Anchor and search bases can be turned on with argument `details = TRUE`. Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign. This happens because the projection is invariant to the sign difference of the bases and so does the index value, however, this creates difficulties for comparing the optimisers since the end bases will be symmetric with respect to the origin. A sign flipping step is conducted to flip the signs of all the bases in one routine if different optimisations finish at opposite places.

Several annotations have been made to help understanding this plot. As mentioned previously, the original basis space is a high dimensional sphere and random bases on the sphere can be generated via the CRAN package \CRANpkg{geozoo} [@geozoo]. Along with the bases recorded during the optimisation and a zero basis, the parameters (centre and radius) of the 2D space can be estimated. PCA is performed to get the first two PC coordinates of all the bases and the centre of the circle is the PC coordinates of the zero matrix, and radius is estimated as the largest distance from the centre to the basis. The theoretical best basis is known for simulated data and can be labelled to compare the how close the end basis found by the optimisers is to the theoretical best. Various aesthetics, i.e. size, alpha and colour, are applicable to emphasize critical elements and adjust for the presentation. For example, anchor points and search points are less important and hence a smaller size and alpha is used; Alpha can also be applied on the interpolation paths to show the start to finish from transparent to opaque.

Figure \ref{fig:toy-pca} shows the PCA plot of SA and PD for a 1D projection problem. Both optimisers find the optimum but PD gets closer. With PCA plot, one can visually appreciate the nature of these two optimisers: PD first evaluates points in a small neighbourhood for a promising direction, while SA evaluates points randomly in the search space to search for the next target. There are dashed lines annotated for SA and it describes the interruption of the interpolation, which will be discussed in details in Section \ref{monotonic}.

@@

Animation is another display to show how the search progresses from start to finish in the space and an  `animate = TRUE` argument is used to enable the animation in the PCA plot. Figure \ref{fig:toy-pca-animated} shows six frames from the animation of the PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information one can learn from this animation is that SA finds its end basis quicker than PD since SA finishes its search in the 5th frame while PD is still making more progression.

@@

As mentioned previously, the original $p \times d$ dimension space can be simulated via randomly generated bases in \CRANpkg{geozoo} [@geozoo] package. While the PCA plot projects the bases from the direction that maximises the variance, the tour plot display of the original high dimensional space from various directions using animation. Figure \ref{fig:toy-tour} shows frames from the tour plot of the same two optimisations in its original 5D space.

@@

In this section, several examples will be presented to show how the diagnostic plots discover something unexpected in projection pursuit optimisation, and guide the implementation of new features.

@@

Random variables with different distributions have been simulated and the distributional form of each variable is presented in Equations \ref{eq:sim-norm} to \ref{eq:sim-x7}. Variable `x1`, `x8`, `x9` and `x10` are normal noise with zero mean and unit variance and `x2` to `x7` are normal mixtures with varied weights and locations. All the variables have been scaled to have an overall unit variance before projection pursuit.

@@

An example of non-monotonic interpolation has been given in Figure \ref{fig:toy-interp}: a path that passes bases with higher index value than the target one. For SAJO, a non-monotonic interpolation is justified since target bases do not necessarily have a higher index value than the current one, while this is not the case for SA. The original trace plot for a 2D projection problem, optimised by SA, is shown on the left panel of Figure \ref{fig:interruption} and one can observe clearly that the non-monotonic interpolation has undermined the optimiser to realise its full potential. Hence, an interruption is implemented to stop at the best basis found in the interpolation. The right panel of Figure \ref{fig:interruption} shows the trace plot after implementing the interruption and while the first two interpolations are identical, the basis at Time 61 has a higher index value than the target in the third interpolations. Rather than starting the next iteration from the target basis on Time 65, SA starts the next iteration at Time 61 on the right panel and reaches a better final basis.

@@

Once the final basis has been found by an optimiser, one may want to push further in the close neighbourhood to see if an even better basis can be found. A polish search takes the final basis of an optimiser as the start of a new guided tour to search for local breakthrough. The polish algorithm is similar to the SA but with three distinctions: 1) a hundred rather than one candidate bases are generated each time in the inner loop; 2) the neighbourhood size is reduced in the inner loop, rather than in the outer loop; and 3) three more termination conditions have been added to ensure the new basis generated is distinguishable from the current one in terms of the distance in the space,  percentage change in the index value, and neighbourhood size:

@@

Figure \ref{fig:polish} presents the trace plot of a 2D projection, optimised by SA and followed by the polish. THe projection by the final basis of each algorithm is also shown. The end basis found by SA reveals the four clusters in the data, but the edges of each cluster are not clean-cut. Polish works with this end basis and further pushes the index value to produce much clear edges of the cluster, especially along the vertical axis.

@@

The holes index function used for all the examples before this section produces a smooth interpolation, while this is not the case for all the indices. A 1D projection function, $I^{nk}(n)$, compares the projected data, $\mathbf{Y}_{n \times 1}$, to a randomly generated normal distribution, $\mathcal{N}_{n \times 1}$, based on the Kolmogorov test. Let $F_{.}(n)$ be the ECDF function, with two possible subscripts $Y$ and $\mathcal{N}$ representing the projected and randomly generated data, and $n$ denoting the number of observation, the Normal Kolmogorov index, $I^{nk}(n)$, is defined as:

$$I^{nk}(n) = \max \left[F_{Y}(n) - F_{\mathcal{N}}(n)\right]$$
With a non-smooth index function, two research questions are raised:

1) whether any optimiser fails to optimise this non-smooth index; and 
2) whether the optimisers can find the global optimum despite the presence of a local optimum

```{r noisy-better-geo, fig.height = 4, fig.cap = "Trace and PCA plot of 1D projection on 5\\-variable dataset, \\code{boa5}, with \\code{norm\\_kol} index, optimised by all three optimiers. Optimiser PD fails to optimise the non-smooth index while although both SA and SAJO success in optimise the index, SAJO takes much longer than SA and finishes off closer to the theoretial best."}
```

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of all three optimisers and as expected, none of the interpolated path is smooth. There is barely any improvement made by PD, indicating its failure in optimising non-smooth index. While SA and SAJO have managed to get close to the index value of the theoretical best, trace plot shows that it takes SAJO much longer to interpolate towards the final basis. This long interpolation path is partially due to the fluctuation in the early iterations, where the SAJO tends to generously accept inferior bases before concentrating near the optimum. The PCA plot shows the interpolation path and search points excluding the last termination iteration. [A sentence for PD]. Comparing the amount of random search done by SA and SAJO, it is surprising that SAJO doesn't carry that many of sampling as SA. Combine the insight from both the trace and PCA plot, one can learn the two different search strategies by SA and SAJO: SA frequently samples in the space and only make a move when an improvement is guaranteed to be made, while SAJO first broadly accepts bases in the space and then starts the extensive sampling in a narrowed subspace. The specific decision of which optimiser to use will depend on the index curve in the basis space but if the basis space is non-smooth, accepting inferior bases at first, as what SAJO has done, can lead to a more efficient search, in terms of the overall number of points evaluated.

```{r kol-result, fig.height = 4, fig.cap="PCA plot of 1D projection on the 6\\-variable data, \\code{boa6}, with normal kolmogorov index. Two optimisers: SA and SAJO and two search sizes: 0.5 and 0.7 are used in the simulation for 20 replicates. Local optimum (x), is presented in this experiment, along with the global optimum (*). The final basis is larger and more opaque than the start basis for each path and color represents whether the global or local optimum is found, after polishing."}
```

The next experiment compares the performance of SA and SAJO when a local maximum exists. Two search neighbourhood sizes: 0.5 and 0.7 are compared to understand how a large search neighbourhood would affect the final basis and the length of the search. Figure \ref{fig:kol-result} shows 80 paths simulated using 20 seeds in the PCA plot, faceted by  optimiser and search size. With SA and a search size of 0.5, despite being the simplest and fastest, the optimiser fails in three instances where the final basis lands neither near the local nor the global optimum. With a larger search size of 0.7, more seeds have found the global maximum. Comparing SA and SAJO for a search size of 0.5, SAJO doesn't seem to improve the final basis found, despite having longer interpolation paths. However, the denser paths near the local maximum is an indicator that SAJO is working hard to examine if there is any other optimum in the basis space but the relative small search size has diminished its ability to reach the global maximum. With a larger search size, almost all the seeds (16 out of 20) have found the global maximum and some final bases are much closer to the theoretical best, as compared to the three other cases. This indicates that SAJO, with a reasonable large search size, is able to overcome the local optimum and optimise close towards the global optimum.

@@

The implementation of this projection has been divided into two packages: the data collection object is implemented in the existing CRAN package \CRANpkg{tourr} [@tourr] while the optimiser diagnostics have been implemented in a new package, \pkg{ferrn}. When a guided tour is run, the users can choose to collect the data by binding `animate_*()` to a variable. Once the data object has been obtained, the \pkg{ferrn} package provides diagnostic plots shown in Section \ref{vis-diag}. The main functions of the \pkg{ferrn} package functionality is listed below.

@@

  - \code{explore\_trace\_search()} produces summary plots in Figure \ref{fig:toy-search}
  - \code{explore\_trace\_interp()}produces trace plots for the interpolation points in Figure \ref{fig:toy-interp}
  - \code{explore\_space\_pca()} produces the PCA plot of projection bases on the reduced space in Figure \ref{fig:toy-pca}. Animated version in Figure \ref{fig:toy-pca-animated} can be turned on via the argument `animate = TRUE`.
  - \code{explore\_space\_tour()} produces animated tour view on the full space of the projection bases in Figure \ref{fig:toy-tour}.

@@

- \code{add\_*()} create ggprotos for different components for the PCA plot
  - \code{add\_anchor()} is a wrapper for plotting anchor bases
  - \code{add\_anno()} is a wrapper for annotating the symmetry of start bases
  - \code{add\_dir\_search()} is a wrapper for plotting the directional search bases with magnified distance
  - \code{add\_end()} is a wrapper for plotting end bases
  - \code{add\_interp()} is a wrapper for plotting the interpolation path
  - \code{add\_interp\_last()} is a wrapper for plotting the last interpolation bases for comparing with target bases when interruption is carried
  - \code{add\_interrupt()} is a wrapper for linking the last interpolation bases with target ones when interruption is carried
  - \code{add\_search()} is a wrapper for plotting search bases
  - \code{add\_space()} is a wrapper for plotting the circular space 
  - \code{add\_start()} is a wrapper for plotting start bases
  - \code{add\_theo()} is a wrapper for plotting theoretical best bases, if applicable

- Utilities
  - \code{theme\_fern()} and \code{format\_label()} for better display of the grid lines and axis formatting
  - \code{clean\_method()} for clean up the name of the optimisers
  - \code{botanical\_palettes} is a collection of colour palettes from Australian native plants Quantitative palettes include daisy, banksia and cherry and sequential palettes contain fern and acacia
  - \code{botanical\_pal()} as the colour interpolator 
  - \code{scale\_color\_*()} and \code{scale\_fill\_*()} for scaling the colour and fill of the plot
